---
title: "Tree <-> Distance matrix"
author: "GAO Zheng"
date: "January 30, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a R-markdown that 

1. simulates binary trees, 
2. calculate the distance matrix of its nodes, and 
3. recoveres the tree structure from the distance matrix.


## simulating binary tree

```{r}
root <- c(1:(1+sample(2,1)))
(root <- c(root,rep(NA,4-length(root))))
names(root) <- c("id","left","right","parent")
(n_nodes <- max(root,na.rm = T))
current <- 2
tree <- list(root)
while (current <= n_nodes) {
  new_children <- sample(0:2, 1, prob = c(0.4,0.4,0.3))
  if (new_children == 2) {
    tree[[current]] <- c(id = current, left = 1+n_nodes, right = 2+n_nodes)
  } else if (new_children == 1) {
    tree[[current]] <- c(id = current, left = 1+n_nodes, right = NA)
  } else {
    tree[[current]] <- c(id = current, left = NA, right = NA)
  }
  n_nodes <- n_nodes + new_children
  current <- current + 1
}

for (current in 1:n_nodes) {
  if (!is.na(tree[[current]]["left"]))
    tree[[tree[[current]]["left"]]]["parent"] = current
  if (!is.na(tree[[current]]["right"]))
  tree[[tree[[current]]["right"]]]["parent"] = current
}

tree
```

## record neighborhood structure

```{r}
A <- diag(n_nodes)
for (i in 1:n_nodes) {
  if (!is.na(tree[[i]]["left"])) {
    A[i, tree[[i]]["left"]] = 1
    A[tree[[i]]["left"], i] = 1
  }
  if (!is.na(tree[[i]]["right"])){
    A[i, tree[[i]]["right"]] = 1
    A[tree[[i]]["right"], i] = 1
  }
}
A
```

## define some helper functions

```{r}
nodes_bt_root <- function(node, root = 1) {
  if (!is.na(root)) {
    x = 0
    if ((root == node) || ((x = nodes_bt_root(node, tree[[root]]["left"])) > 0)
        || ((x = nodes_bt_root(node, tree[[root]]["right"])) > 0)) {
      return(x + 1)
    }
    return(0)
  }
  return(0)
}

dist_to_root <- function(node) {
  nodes_bt_root(node) - 1
}

trace_to_root <- function(node) {
  parent <- tree[[node]]["parent"]
  names(parent) <- NULL
  if (!is.na(node)) {
    return(c(trace_to_root(parent),node))
  }
}

lowest_comm_ascestor <- function(node1, node2) {
  trace_n1 <- trace_to_root(node1)
  trace_n2 <- trace_to_root(node2)
  max_len <- max(length(trace_n1),length(trace_n2))
  trace_n1 <- c(trace_n1, rep(0, max_len - length(trace_n1)), 1)
  trace_n2 <- c(trace_n2, rep(0, max_len - length(trace_n2)), 2)
  trace_n1[which((trace_n1 - trace_n2) != 0)[1] - 1]
}

dist_bt_nodes <- function(node1, node2) {
  dist_to_root(node1) + dist_to_root(node2) - 
    2 * dist_to_root(lowest_comm_ascestor(node1,node2))
}
```

## calculate the distance matrix

```{r}
distance_matrix <- function() {
  n_nodes <- length(tree)
  D <- matrix(Inf, n_nodes, n_nodes)
  for (i in 1:n_nodes) {
    for (j in 1:n_nodes) {
      D[i,j] <- dist_bt_nodes(i,j)
    }
  }
  return(D)
}

(D <- distance_matrix())
```

## tree recovery

**ALGO**:

1. construct a fully connected graph with edge weights d(i,j).
2. remove all edges (i,j) if d(i,j) = d(i,k) + d(k,j) for some k != i,j.

**Pf**:

if d(i,j) = d(i,k) + d(k,j) for some k, 
then there must not be an edge between i and j (since tree is acyclic, the path is unique)
So every removed edge is not in the tree

On the other hand,
if an edge (i,j) is not in the original tree,
then there necessarily exists a node k in the shortest path from i to j,
i.e., d(i,j) = d(i,k) + d(k,j) for some k,
So edges that are not in the tree are removed. 


```{r}
recover_tree <- function(D) {
  n_nodes <- nrow(D)
  A_recovered <- matrix(1,n_nodes,n_nodes)
  for (i in 1:n_nodes) {
    for (j in 1:n_nodes) {
      for (k in 1:n_nodes) {
        if ((i != j) && 
            (k != i) && (k != j) &&
            (D[i,j] >= D[i,k] + D[k,j]))
          A_recovered[i,j] <- 0
      }
    }
  }
  return(A_recovered)
}

(A_recovered <- recover_tree(D))

A == A_recovered
```
